\documentclass[11pt]{article}
\usepackage[margin=1.8cm]{geometry}
\usepackage{tikz-cd}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{enumitem}
\setlist{nosep}
\pagestyle{empty}

\newcommand{\Adm}{\mathrm{Adm}}
\newcommand{\Bool}{\mathbf{2}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Kl}{\mathrm{Kl}}
\newcommand{\StateT}{\mathrm{StateT}}
\newcommand{\MC}{\mathrm{MaterialClass}}
\newcommand{\TS}{\mathrm{ThermoState}}
\newcommand{\DIB}{\mathcal{D}\!\mathcal{I}\!\mathcal{B}}

\begin{document}

\begin{center}
{\Large\bfseries UMST Categorical Structure}\\[4pt]
{\small Commuting diagrams for the thermodynamic gate kernel}
\end{center}

\medskip

\noindent\textbf{1.\enspace Gate as Natural Transformation.}\enspace
The gate $G$ is natural in the material class: for every functor
$F_m : \MC \to \TS$ constructing a thermodynamic state from mix parameters,
the gate verdict is independent of the material path.

\[
\begin{tikzcd}[column sep=3.5em, row sep=3em]
\MC \times \MC
  \arrow[r, "F_m \times F_m"]
  \arrow[d, "f \times f"']
& \TS \times \TS
  \arrow[r, "G"]
  \arrow[d, "F(f) \times F(f)"]
& \Bool
  \arrow[d, equal]
\\
\MC' \times \MC'
  \arrow[r, "F_{m'} \times F_{m'}"']
& \TS \times \TS
  \arrow[r, "G"']
& \Bool
\end{tikzcd}
\]

\noindent The square on the left commutes by functoriality of state construction;
the square on the right expresses naturality of $G$:
$G \circ (F(f) \times F(f)) = G$.
The gate checks only thermodynamic quantities, not material identity.

\medskip

\noindent\textbf{2.\enspace DIB Kleisli Triangle.}\enspace
Discovery ($D$), Invention ($I$), and Build ($B$) are morphisms in the
Kleisli category $\Kl(\StateT\;\TS\;\mathrm{IO})$.
Their composition yields an admissibility-preserving loop:

\[
\begin{tikzcd}[column sep=3em, row sep=2.8em]
& \TS
  \arrow[dr, "I"]
\\
\TS
  \arrow[ur, "D"]
  \arrow[rr, "B \circ I \circ D"', bend right=15]
& & \TS
  \arrow[ll, "B", bend right=15]
\end{tikzcd}
\qquad
\begin{aligned}
D &: \TS \to \Kl(\TS) \\
I &: \TS \to \Kl(\TS) \\
B &: \TS \to \Kl(\TS) \\
\DIB &= B \circ_{\Kl} I \circ_{\Kl} D
\end{aligned}
\]

\noindent Kleisli composition guarantees that if each phase individually
preserves admissibility (i.e., the gate accepts each transition), then
the full $\DIB$ loop preserves it.

\medskip

\noindent\textbf{3.\enspace Monoidal Mass Conservation.}\enspace
The state space $\TS$ carries a monoidal structure $(\TS, \oplus, \mathbf{0})$
where $\oplus$ is the tensor product of independent subsystems (e.g., paste
+ aggregate). Mass conservation is a monoidal constraint:

\[
\begin{tikzcd}[column sep=4em, row sep=2.8em]
\TS \otimes \TS
  \arrow[r, "\rho \otimes \rho"]
  \arrow[d, "\oplus"']
& \mathbb{Q} \times \mathbb{Q}
  \arrow[d, "{+}"]
\\
\TS
  \arrow[r, "\rho"']
& \mathbb{Q}
\end{tikzcd}
\qquad
|\rho(s_1 \oplus s_2) - (\rho(s_1) + \rho(s_2))| < \delta
\]

\noindent The density functional $\rho$ is a monoidal morphism up to
tolerance $\delta$, ensuring that composing subsystems cannot create or
destroy mass beyond measurement precision.

\medskip

\noindent\textbf{Summary.}\enspace
Three structures proved in the Agda layer:
\emph{(i)}~naturality of the gate,
\emph{(ii)}~Kleisli composability of DIB,
\emph{(iii)}~monoidal mass conservation.
Together they establish that the thermodynamic gate is a
\emph{categorically well-founded} decision procedure over physically
grounded invariants.

\end{document}
